Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-02 article-models-tt.tex
filled in all the proofs

\begin{defn}
For $\jhom{\Gamma}{A}{B}{f}$ and consider a term $\jterm{{\Gamma}{B}}{Q}{g}$.
We define
\begin{align*}
\jfamdefn*{{\Gamma}{A}}{\jhomcomp{f}{Q}}{\subst{f}{\ctxwk{A}{Q}}}\\
\jtermdefn*{{\Gamma}{A}}{\jhomcomp{f}{Q}}{\jhomcomp{f}{g}}{\subst{f}{\ctxwk{A}{g}}}.
\end{align*}
Likewise, when we have a family $\jfam{{{\Gamma}{B}}{Q}}{R}$ and a term
$\jterm{{{\Gamma}{B}}{Q}}{R}{h}$, we define
\begin{align*}
\jfamdefn*{{{\Gamma}{A}}{\jhomcomp{f}{Q}}}{\jhomcomp{f}{R}}{\subst{f}{\ctxwk{A}{R}}}\\
\jtermdefn*{{{\Gamma}{A}}{\jhomcomp{f}{Q}}}{\jhomcomp{f}{R}}{\jhomcomp{f}{h}}{\subst{f}{\ctxwk{A}{h}}}.
\end{align*}
\end{defn}

@@

\begin{lem}
We have the following inference rules about the situation where something is
substituted by a composition:
\begin{align*}
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jfam{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}}
  {\jfameq{{\Gamma}{A}}{\subst{\jhomcomp{f}{g}}{R}}{\subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{R}}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jterm{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}{h}}
  {\jfameq{{\Gamma}{A}}{\subst{\jhomcomp{f}{g}}{h}}{\subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{h}}}}}
\end{align*}
In particular, when $\jhom{\Gamma}{C}{D}{h}$, we have
\begin{equation*}
\jhomeq{\Gamma}{A}{D}{\jhomcomp{\jhomcomp{f}{g}}{h}}{\jhomcomp{f}{\jhomcomp{g}{h}}}.
\end{equation*}
\end{lem}

\begin{proof}
Consider family morphisms $\jhom{\Gamma}{A}{B}{f}$ and $\jhom{\Gamma}{B}{C}{g}$
and a family $\jfam{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}$. Then we have the judgmental
equalities
\begin{align*}
\subst{\jhomcomp{f}{g}}{R} & \jdeq \subst{{f}{\ctxwk{A}{g}}}{R}\\
& \jdeq \subst{{f}{\ctxwk{A}{g}}}{\subst{f}{\ctxwk{{A}{B}}{R}}}\\
& \jdeq \subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{R}}}
\end{align*}
\end{proof}

\begin{lem}
We have the following inference rules about the compatibility of composition with
weakening:
\begin{align*}
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jfam{{\Gamma}{A}}{P}}
  {\jhomeq{\Gamma}{{A}{P}}{C}{\ctxwk{P}{\jhomcomp{f}{g}}}{\jhomcomp{\ctxwk{P}{f}}{g}}}\\
& \inference
  {\jterm{\Gamma}{B}{y}
   \jhom{\Gamma}{B}{C}{g}}
  {\jhomeq{\Gamma}{A}{C}{\jhomcomp{\ctxwk{A}{y}}{g}}{\ctxwk{A}{\subst{y}{g}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jterm{\Gamma}{C}{z}}
  {\jhomeq{\Gamma}{A}{C}{\jhomcomp{f}{\ctxwk{B}{z}}}{\ctxwk{A}{z}}}
\end{align*}
\end{lem}

\begin{lem}
We have the following inference rules about the compatibility of composition with
substitution:
\begin{align*}
& \inference
  {\jhom{{\Gamma}{A}}{P}{Q}{f}
   \jhom{{\Gamma}{A}}{Q}{R}{g}
   \jterm{\Gamma}{A}{x}}
  {\jhomeq{\Gamma}{\subst{x}{P}}{\subst{x}{R}}{\subst{x}{\jhomcomp{f}{g}}}{\jhomcomp{\subst{x}{f}}{\subst{x}{g}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jterm{\Gamma}{A}{x}}
  {\jtermeq{\Gamma}{C}{\subst{x}{\jhomcomp{f}{g}}}{\subst{{x}{f}}{g}}}
\end{align*}
\end{lem}

@@

\begin{lem}
For any $\jfam{\Gamma}{A}$ and $\jfam{{\Gamma}{A}}{P}$ we have
\begin{equation}\label{idfunc-ext-comp}
\inference
  {\jfam{\Gamma}{A}
   \jfam{{\Gamma}{A}}{P}}
  {\jtermeq{{\Gamma}{{A}{P}}}{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}{\idfunc[\ctxext{A}{P}]}{\ctxext{\cprojfstf{A}{P}}{\cprojsndf{A}{P}}}}
\end{equation}
\end{lem}

\begin{proof}
Consider the families $\jfam{\Gamma}{A}$ and $\jfam{{\Gamma}{A}}{P}$. Then
we have the judgmental equalities
\begin{align*}
\ctxext{\cprojfstf{A}{P}}{\cprojsndf{A}{P}}
& \jdeq \subst{\cprojsndf{A}{P}}{{\cprojfstf{A}{P}}{\idfunc[\ctxext{\ctxwk{\ctxext{A}{P}}{A}}{\ctxwk{\ctxext{A}{P}}{P}}]}}\\
& \jdeq \subst{\cprojsndf{A}{P}}{{\cprojfstf{A}{P}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}\\
& \jdeq \subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}\\
& \jdeq \subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}\\
& \jdeq \subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{\idfunc[\ctxext{A}{P}]}}}}\\
& \jdeq \subst{\idfunc[P]}{\ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{\idfunc[\ctxext{A}{P}]}}}}\\
& \jdeq \subst{\idfunc[A]}{\ctxwk{A}{\idfunc[\ctxext{A}{P}]}}\\
& \jdeq \idfunc[\ctxext{A}{P}].
\end{align*}
\end{proof}

>>>>

\begin{defn}
For $\jhom{\Gamma}{A}{B}{f}$ and consider a term $\jterm{{\Gamma}{B}}{Q}{g}$.
We define
\begin{align*}
\jfamdefn*{{\Gamma}{A}}{\jcomp{f}{Q}}{\subst{f}{\ctxwk{A}{Q}}}\\
\jtermdefn*{{\Gamma}{A}}{\jcomp{f}{Q}}{\jcomp{f}{g}}{\subst{f}{\ctxwk{A}{g}}}.
\end{align*}
Likewise, when we have a family $\jfam{{{\Gamma}{B}}{Q}}{R}$ and a term
$\jterm{{{\Gamma}{B}}{Q}}{R}{h}$, we define
\begin{align*}
\jfamdefn*{{{\Gamma}{A}}{\jcomp{f}{Q}}}{\jcomp{f}{R}}{\subst{f}{\ctxwk{A}{R}}}\\
\jtermdefn*{{{\Gamma}{A}}{\jcomp{f}{Q}}}{\jcomp{f}{R}}{\jcomp{f}{h}}{\subst{f}{\ctxwk{A}{h}}}.
\end{align*}
\end{defn}

@@

\begin{lem}
We have the following inference rules about the situation where something is
substituted by a composition:
\begin{align*}
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jfam{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}}
  {\jfameq
    {{\Gamma}{A}}
    {\subst{\jcomp{f}{g}}{R}}
    {\subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{R}}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jterm{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}{h}}
  {\jfameq
    {{\Gamma}{A}}
    {\subst{\jcomp{f}{g}}{h}}
    {\subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{h}}}}}
\end{align*}
We also have the following related inference rules, asserting that composition
is strictly associative:
\begin{align*}
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jfam{{\Gamma}{C}}{R}}
  {\jfameq
    {{\Gamma}{A}}
    {\jcomp{{f}{g}}{R}}
    {\jcomp{f}{{g}{R}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jterm{{\Gamma}{C}}{R}{h}}
  {\jtermeq
    {{\Gamma}{A}}
    {\jcomp{{f}{g}}{R}}
    {\jcomp{{f}{g}}{h}}
    {\jcomp{f}{{g}{h}}}}
\end{align*}
\end{lem}

\begin{proof}
Consider family morphisms $\jhom{\Gamma}{A}{B}{f}$ and $\jhom{\Gamma}{B}{C}{g}$
and a family $\jfam{{{\Gamma}{A}}{\ctxwk{A}{C}}}{R}$. Then we have the judgmental
equalities
\begin{align*}
\subst{\jcomp{f}{g}}{R} & \jdeq \subst{{f}{\ctxwk{A}{g}}}{R}\\
& \jdeq \subst{{f}{\ctxwk{A}{g}}}{\subst{f}{\ctxwk{{A}{B}}{R}}}\\
& \jdeq \subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{R}}}
\end{align*}
The proof that 
$\subst{\jcomp{f}{g}}{h}\jdeq\subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{h}}}$
is similar.

Now suppose that $\jfam{{\Gamma}{C}}{R}$ instead. Then we have
\begin{align*}
\jcomp{{f}{g}}{R} & \jdeq \subst{\jcomp{f}{g}}{\ctxwk{A}{R}}\\
& \jdeq \subst{{f}{\ctxwk{A}{g}}}{\ctxwk{A}{R}}\\
& \jdeq \subst{f}{{\ctxwk{A}{g}}{\ctxwk{{A}{B}}{{A}{R}}}}\\
& \jdeq \subst{f}{{\ctxwk{A}{g}}{\ctxwk{A}{{B}{R}}}}\\
& \jdeq \subst{f}{\ctxwk{A}{\subst{g}{\ctxwk{B}{R}}}}\\
& \jdeq \subst{f}{\ctxwk{A}{\jcomp{g}{R}}}\\
& \jdeq \jcomp{f}{{g}{R}}.
\end{align*}
Again, the proof is similar for terms $h$ of $R$ in context $\ctxext{\Gamma}{C}$.
\end{proof}

\begin{lem}
We have the following inference rules about the compatibility of composition with
weakening:
\begin{align*}
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jfam{{\Gamma}{A}}{P}}
  {\jhomeq{\Gamma}{{A}{P}}{C}
    {\ctxwk{P}{\jcomp{f}{g}}}
    {\jcomp{\ctxwk{P}{f}}{g}}}\\
& \inference
  {\jterm{\Gamma}{B}{y}
   \jhom{\Gamma}{B}{C}{g}}
  {\jhomeq{\Gamma}{A}{C}
    {\jcomp{\ctxwk{A}{y}}{g}}
    {\ctxwk{A}{\subst{y}{g}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jterm{\Gamma}{C}{z}}
  {\jhomeq{\Gamma}{A}{C}
    {\jcomp{f}{\ctxwk{B}{z}}}
    {\ctxwk{A}{z}}}
\end{align*}
\end{lem}

\begin{proof}
Let $\jhom{\Gamma}{A}{B}{f}$, $\jhom{\Gamma}{B}{C}{g}$ and $\jfam{{\Gamma}{A}}{P}$.
Then we have the judgmental equalities
\begin{align*}
\ctxwk{P}{\jcomp{f}{g}} 
& \jdeq \ctxwk{P}{\subst{f}{\ctxwk{A}{g}}}\\
& \jdeq \subst{\ctxwk{P}{f}}{\ctxwk{P}{{A}{g}}}\\
& \jdeq \subst{\ctxwk{P}{f}}{\ctxwk{\ctxext{A}{P}}{g}}\\
& \jdeq \jcomp{\ctxwk{P}{f}}{g}.
\end{align*}
Now let $\jterm{\Gamma}{B}{y}$ and $\jhom{\Gamma}{B}{C}{g}$. Then we have the
judgmental equalities
\begin{align*}
\jcomp{\ctxwk{A}{y}}{g}
& \jdeq \subst{\ctxwk{A}{y}}{\ctxwk{A}{g}}\\
& \jdeq \ctxwk{A}{\subst{y}{g}}.
\end{align*}
For the third assertion, let $\jhom{\Gamma}{A}{B}{f}$ and $\jterm{\Gamma}{C}{z}$.
Then we have the judgmental equalities
\begin{align*}
\jcomp{f}{\ctxwk{B}{z}} 
& \jdeq \subst{f}{\ctxwk{A}{{B}{z}}}\\
& \jdeq \subst{f}{\ctxwk{{A}{B}}{{A}{z}}}\\
& \jdeq \ctxwk{A}{z}.
\end{align*}
\end{proof}

\begin{lem}
We have the following inference rules about the compatibility of composition with
substitution:
\begin{align*}
& \inference
  {\jhom{{\Gamma}{A}}{P}{Q}{f}
   \jhom{{\Gamma}{A}}{Q}{R}{g}
   \jterm{\Gamma}{A}{x}}
  {\jhomeq{\Gamma}{\subst{x}{P}}{\subst{x}{R}}
    {\subst{x}{\jcomp{f}{g}}}
    {\jcomp{\subst{x}{f}}{\subst{x}{g}}}}\\
& \inference
  {\jhom{\Gamma}{A}{B}{f}
   \jhom{\Gamma}{B}{C}{g}
   \jterm{\Gamma}{A}{x}}
  {\jtermeq{\Gamma}{C}
    {\subst{x}{\jcomp{f}{g}}}
    {\subst{{x}{f}}{g}}}
\end{align*}
\end{lem}

\begin{proof}
Let $\jhom{{\Gamma}{A}}{P}{Q}{f}$, $\jhom{{\Gamma}{A}}{Q}{R}{g}$ and $\jterm{\Gamma}{A}{x}$.
Then we have the judgmental equalities
\begin{align*}
\subst{x}{\jcomp{f}{g}}
& \jdeq \subst{x}{{f}{\ctxwk{P}{g}}}\\
& \jdeq \subst{{x}{f}}{{x}{\ctxwk{P}{g}}}\\
& \jdeq \subst{{x}{f}}{\ctxwk{\subst{x}{P}}{\subst{x}{g}}}\\
& \jdeq \jcomp{\subst{x}{f}}{\subst{x}{g}}.
\end{align*}
Now let $\jhom{\Gamma}{A}{B}{f}$, $\jhom{\Gamma}{B}{C}{g}$ and $\jterm{\Gamma}{A}{x}$.
Then we have the judgmental equalities
\begin{align*}
\subst{x}{\jcomp{f}{g}}
& \jdeq \subst{x}{{f}{\ctxwk{A}{g}}}\\
& \jdeq \subst{{x}{f}}{{x}{\ctxwk{A}{g}}}\\
& \jdeq \subst{{x}{f}}{g}.
\end{align*}
\end{proof}

@@

\begin{lem}
For any $\jfam{\Gamma}{A}$ and $\jfam{{\Gamma}{A}}{P}$ we have
\begin{equation}\label{idfunc-ext-comp}
\inference
  {\jfam{\Gamma}{A}
   \jfam{{\Gamma}{A}}{P}}
  {\jtermeq
    {{\Gamma}{{A}{P}}}
    {\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}
    {\idfunc[\ctxext{A}{P}]}
    {\ctxext{\cprojfstf{A}{P}}{\cprojsndf{A}{P}}}}
\end{equation}
\end{lem}

\begin{proof}
Consider the families $\jfam{\Gamma}{A}$ and $\jfam{{\Gamma}{A}}{P}$. Then
we have the judgmental equalities
\begin{align*}
\ctxext{\cprojfstf{A}{P}}{\cprojsndf{A}{P}}
& \jdeq \subst{\cprojsndf{A}{P}}{{\cprojfstf{A}{P}}{\idfunc[\ctxext{\ctxwk{\ctxext{A}{P}}{A}}{\ctxwk{\ctxext{A}{P}}{P}}]}}\\
& \jdeq \subst
          {\cprojsndf{A}{P}}
          { {\cprojfstf{A}{P}}
            {\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}\\
& \jdeq \subst
          {\idfunc[P]}
          { {\ctxwk{P}{\idfunc[A]}}
            {\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}\\
& \jdeq \subst
          {\idfunc[P]}
          { {\ctxwk{P}{\idfunc[A]}}
            {\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}\\
& \jdeq \subst
          {\idfunc[P]}
          { {\ctxwk{P}{\idfunc[A]}}
            {\ctxwk{P}{{A}{\idfunc[\ctxext{A}{P}]}}}}\\
& \jdeq \subst
          {\idfunc[P]}
          {\ctxwk
            {P}
            {\subst
              {\idfunc[A]}
              {\ctxwk{A}{\idfunc[\ctxext{A}{P}]}}}}\\
& \jdeq \subst
          {\idfunc[A]}
          {\ctxwk{A}{\idfunc[\ctxext{A}{P}]}}\\
& \jdeq \idfunc[\ctxext{A}{P}].
\end{align*}
\end{proof}
